---
title: "Exploring the Real Estate Market in Upstate New York"
output: word_document
---

In this HW, you will explore data collected for houses sold
in Saratoga, New York in 2007.  We will explore house prices, size of the
house, and a few other variables such as the number of bedrooms and
the number of fireplaces. 

The main goal of this HW is to give you practice working with 
differnt data types, subsetting, and creating vectors.  A side 
benefit will be that you will gain practice thinking with data 
as you carry out your computations. 
The analysis of the data includes exmaning univariate distributions,
exploring relationships between variables, and plotting.

-----------

## Reading the Data Into R

Your first task is to read the data into R.
The data are located on the web at 

http://www.stat.berkeley.edu/users/nolan/data/stat133/Saratoga.txt

- Visit the Web site and exmaine the layout of the data. 
Do not download it to your computer. 

    * Do the data have a header containing the variable names? Yes
    * Are the values for an observation separated by a comma, blank, or
tab? Tab

- Read the documentation for read.csv() and read.table() and examine
the parameters that are used to specify whether the data have a header
and how the values in a row are separated. Notice that these functions
are similar, but they have different default values for these 
parameters.

- Use read.csv() to read the data from the Web into a data frame 
called *house*. (Remember to place the URL within quotes and 
pass it as an argument to the url() function.)

- Use read.table() to read the data from the Web into a data frame
called *house2*.

- Check that these two data frames are identical. Use the function
identical() to do this.

```{r}
house <- read.csv(url("http://www.stat.berkeley.edu/users/nolan/data/stat133/Saratoga.txt"), sep = "")
house2 <- read.table("http://www.stat.berkeley.edu/users/nolan/data/stat133/Saratoga.txt", header = T)
identical(house,house2)
```

-----------------

## Examine Price 

We start by exploring house price. 
Make summary statistics of price. 
Remember that you can access a variable within a data frame 
with the $ sign.


```{r}
summary(house2$Price)
```

It's a bit surprising to see that the lowest price is only $5,000.
Also notice that the mean price is about $25,000 more than the median
price, which indicates that the distribution of price may be skewed
right, possibly by having a long right tail.

Make histogram of house price. No need to give fancy labels
at this point; we're just exploring the data.
Be sure to specify enough bins to see the shape of the data.
Indeed, we do find that the prices are skewed to the right 
with relatively few houses priced above $400K. 
That skewness explains why the mean is higher than the median. 
The few houses priced between $400K and $800K have pulled it away 
from the typical price. 

```{r}
hist(house2$Price, 
     breaks = 30, 
     main = "Histogram of House Price",
     xlab = "Price")
```
Does a log transformation make the price distribution more symmetric?
Now we see those unusually small values cropping up on the left.
However the rest of the distribution looks symmetric and unimodal.

```{r}
hist(log(house2$Price),
     breaks = 30, 
     main = "Histogram of Log of House Price",
     xlab = "Log of Price")
```

Let's use subsetting to dig a little deeper into the ussue of the
unusually inexpensive houses. We will answer the questions:

- How many are there?  5
- Do they have unusual values for the other variables? Yes

To answer the first of these questions, create a logical vector
that indicates if the log of the house price is under 10.5.
Assign this logical vector to the variable named `cheap'.

```{r}
cheap <- log(house2$Price) < 10.5
```

How many houses are what we call cheap?

```{r}
length(house2$Price[cheap == TRUE])
```

Use subsetting to print the values of all of the variables 
for the cheap houses.
# logical into variable

```{r}
house[(log(house2$Price) < 10.5)==T,]
```

These houses do not appear to have unusual values for the other
variables. 

Use the variable cheap to subset the data frame house
and remove these rows from the data frame.
Call the new data frame `house'; that is, replace the existing
data frame with the new one.

```{r}
house2[house2$cheap,]
```

## Explore Living Area

Next we explore the size of the house, i.e., the living area.
Do you find a similar issue with the living area? 
Examine the summary statistics for living area and make 
a histogram of the values for living area and another
histogra for the logarithm of these values. 

```{r}
summary(house2$Living.Area)
hist(house$Living.Area, main = "Histogram of Living Area", xlab = "Living Area")
hist(log(house$Living.Area), main = "Histogram of Log of Living Area", xlab = "Log of Living Area")
```

We again find that the distirbution is skewed to the right.
The log transformation of the values has a less peaked
distribution that is roughly symmetric, but there appears to
be an unusually large number of values at about 6.8.
What value does this correspond to in the original units of measurement?


```{r}
exp(6.8)
```

-----------

## Transforming Variables into Factors

We next transform the number of bedrooms into a factor vector, i.e., 
into a categorical variable.  We want to collapse the number of
bedrooms into four categories, 2 or fewer, 3, 4, and 5 or more.

To do this, follow the following approach:

- Assign the variable *Bedrooms* in *house* to a vector called
*BR*.

- Use subsetting to set all of the values in *BR* that are greater than
5 to 5.

- Use the same approach to set all of values in *BR* that are under 2 
to 2.

- Use the *factor()* function to convert *BR* to a factor vector.
Read carefully about the levels and labels parameters to the function.
For the categories, use the following strings: "2- BR", "3 BR", "4 BR",
and "5+ BR". Assign the factor to *BR*, i.e., over write the 
numeric *BR* with the factor *BR*. It will be helpful to use the ":"
function and the *c()* function to create the arguments to the
function call.

```{r}
BR <- house$Bedrooms
BR[house2$Bedrooms > 5] <- 5
BR[house2$Bedrooms < 2] <- 2
BR <- factor(BR)
```

How does a summary of *BR* look different than a summary of *Bedrooms*?

```{r}
summary(BR)
summary(house2$Bedrooms)
```

-----------

## Plotting Price against Lining Area

We will now make a plot of price against living area 
(both on the log scale)
and color the plotting symbols according to the number of
bedrooms in the house.
We will use the new vector *BR* to determine the color.

Begin by making a vector of 3 colors. Use the colors called
aquamarine3, darkgoldenrod2, coral2, and mediumorchid3, in that 
order. Call this vector, *myColors*.

```{r}
myColors <- c("aquamarine3","darkgoldenrod2","coral2", "mediumorchid3")
plot(x = log(house2$Living.Area), 
     y = log(house2$Price), 
     col = myColors, 
     main = "Log of price vs log of Living area",
     xlab = "Log of Living Area",
     ylab = "Log of Price")
```

Next make vector of colors that matches the length of *BR*, and
where a value is "2- BR", the color is aquamarine3, where a value 
is "3 BR", the color is darkgoldenrod2, etc.  Use subsetting
by position of the vector *myColors* to do this.
Call this new vector, *BRcolors*.

```{r}
myColors[BR == 2] <- 'aquamarine3'
myColors[BR == 3] <- 'darkgoldenrod2'
myColors[BR == 4] <- 'coral2'
myColors[BR == 5] <- "mediumorchid3"
BRcolors <- myColors
```

Lastly, we make the scatterplot as follows. Once you have created
all of the necessary variables, i.e., BR, BRcolors and myColors,
change the eval=FALSE in the R chunk below to eval=TRUE.

```{r, eval= T}
plot( Price ~ Living.Area, data = house, log = "xy", main = "", 
      xlab = "Living Area (log sq ft)", ylab = "Price (log $)",
      col = BRcolors, pch = 19, cex = 0.4)

legend("bottomright", fill = myColors, legend = levels(BR),
       title = "# Bedrooms", cex = 0.75)
```

Don't worry about the various arguments to the function call. 
We will be learning about them this week.

We see in the plot that price and living area have a roughly linear
relationship and that houses with more bedrooms tend to command
a higher price. This makes sense. 
It may be surprising the number of bedrooms is not more 
highly correlated with price and size of the house.
-----------

### Number of Fireplaces

The original variable *Fireplaces* is the actual number of fireplaces
in the house. 
Follow the approach in the previous section to create a new variable
called *FP* that is a factor with just two levels: 
"None","At least 1". 


```{r}
FP <- factor(c("None", "At least 1"))
```

Create a new variable that contains the price per square foot of 
the house.
Make a plot of price per square foot against price (determine
if they should be on the log scale)
and color the plotting symbols according to whether or not there
is a fireplace in the house. To do this, create a vector of colors 
called FPcolors.

```{r}
p <- (house2$Price)/(house2$Living.Area)
FPcolors <- house$Fireplaces == T
plot(p ~ house2$Price, data = house, log = "xy", main = "", xlab = "Price per sq foot", ylab = "Price", col = FPcolors, pch = 19, cex = 0.4)
legend("bottomright", fill = FPcolors, legend = levels(FP),
title = "number of fireplaces", cex = 0.75)
```
plot(p ~ house2$Price, data = house, log = "xy", main = "", xlab = "Price per sq foot", ylab = "Price",  pch = 19, cex = 0.4)
legend("bottomright", legend = levels(FP),
title = "number of fireplaces", cex = 0.75)

Are houses with fireplaces generally more expensive than houses without?
What is the relationship between price per square foot and price?
Can you explain why you might have this shape of relationship?
No. As you can the plot below, more samples are drawn and some data shows the an increase in price/sq foot yields an increase in price. The price per square foot and price is directionally proportional. This shape of relationship explains if a house is more spacious, the more of the house owner has to pay.

```{r}
plot(p ~ house2$Price, data = house, log = "xy", main = "", xlab = "Price per sq foot", ylab = "Price",  pch = 19, cex = 0.4)
legend("bottomright", legend = levels(FP),
title = "number of houses", cex = 0.75)
```
