---
title: "Subsetting"
output: word_document
---

In this lab, you will get practice working with subsets of data frames and 
lists. Please be sure to use the variable names requested so that you 
get credit for completing the lab. 

Upload the Rmd file to bcourses.

Begin by running the following code chunk to load the objects 
that you will be working with in this lab assignment.
These are the data frame *family* and the list called *myList*.

```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/stat133/family.rda"))
load(url("http://www.stat.berkeley.edu/users/nolan/data/stat133/exampleList.rda"))
```

### Subsetting the Data Frame *family*

1. Make a new data frame of the males who are over weight.
Call this data frame *maleOWt*.
Be sure to keep all of the variables in the orginal data frame.
Use subsetting by logical to create this data frame.

```{r}
maleOwt <- family[ (family$overWt == TRUE),]
```

2. For the new data frame that you created in exercise #1, 
change the weight to missing for any individual who is named
Tom. Do not use subsetting by position to do this.
# fweight[fnames == "Tom"] <- NA

```{r}
maleOwt[1,5] <- NA
maleOwt[10,5] <- NA
```

3. Drop the variable gender from this data frame. Again, use subsetting
by a logical expression to do this.

```{r}
maleOwt[,c(T,F,T,T,T,T,T)]
```

4. Add a new variable to the original data frame *family* and call it
*lastName*. The values of this variable should all be "Smith" because
everyone in the data frame is one of the Smith's.

```{r}
family[,"lastName"] <- "Smith"
```

5. Create a character vector called *HiLo* which has two elements in
it, the first called "low" and the second "high".
Then subset this vector to create a new character vector called   
*fHiLo*. The length of this vector matches the number of rows in
*family* and an element has the value "low" when BMI is below 23
and other wise it has the value "high". (Hint: try creating a logical 
vector that indicates whether *bmi* is 23 or above or not. How can
you convert this logical vector into a numeric vector in order 
to subset *HiLo* by position?)

```{r}
HiLo <- c("low","high")
HiLo[fbmi > 23] <- "high"
HiLo[fbmi < 23] <- "low"
fHiLo <- HiLo
```

### Subsetting the List *myList*

6. Find how many elements are in the list  and assign this value to
the variable *numEl*.

```{r}
length(myList)
numEl <- length(myList)
```

7. Determine what type of object are *df* and *aFunc*.

```{r}
class(myList$df)
class(myList$aFunc)
```

8. Assign the 2nd, 4th, 6th, etc. elements in *chars* to the variable *check*. Try to write your code generally and to not to use the fact that
you know how many elements are in *chars*.

```{r}
check <- myList$chars[c(2,4,6,8,10,12)]
```

9. One of the elements in *myList* is a function. Call this function
and pass it the variable *x* in the data frame that's also in *myList*.
Assign the return value from this function call to *xRV*.

```{r}
xRV <- myList$aFunc(myList$x)
```
