---
Name: Ryan Chui
Class: Stat 133
title: "Introduction to Plotting in R"
output: word_document
---

This plotting exercise introduces many of the features in the R graphics functions. 
There are many more features to investigate.
Documentation can be found by calling plot.default()
and also by asking for help with the par function, i.e. ?par
In addition, you might want to visit some of the R visualization
web sites to learn more about making beautiful plots.
These include:

- http://rdataviz.wordpress.com/
- http://gallery.r-enthusiasts.com/
- http://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html
- http://blog.revolutionanalytics.com/2009/01/10-tips-for-making-your-r-graphics-look-their-best.html

### Reading the Data into R

We begin by loading in our data and determining the size and shape and names 
of the variables in the data frame:

```{r}
chips = 
  read.table("http://www.stat.berkeley.edu/users/nolan/data/chip04.txt", 
             header = TRUE)

names(chips)
dim(chips)
head(chips)
```

### A First Attempt to Plot

We want to look at how *Transistors* has changed over the years.
Rather than plot points, we make a line plot that connects the points.

```{r}
plot(x = chips$Date, y = chips$Transistors, type = "l")
```

Make the following changes to the plot:

* Use the formula language to express the relationship that you wanted plotted
* Make the line thicker
* Plot the number of transistors on a log scale

Put your modified code below.

```{r}
plot(log(chips$Transistors)~log(chips$Date), type = "l", lwd = 3)
```

### Adding Context to the Plot

We next add axes labels and a title to the plot
We store the labels in variables to make it easier if we need to make the plot again.

```{r}
titleP = "Intel Microprocessor"
ylabel = "ADD YOUR LABEL"
xlabel = "Year"
```


### Transforming the Data

We transform the data so that we examine relative growth
This way, we can add multiple line graphs to the same canvas
because they will be unitless. Create two other transformed 
variables for clockspeed and mips. Call these *cspd* and *mps*
and add them to the new data frame *chipsT*.

```{r}
cspd = chips$ClockSpeed
mps = chips$MIPS
trans = chips$Transistors/chips$Transistors[1]
chipsT = data.frame(date = chips$Date,
                    trans = chips$Transistors/chips$Transistors[1], 
                    cspd, mps
                    )
```

Before we make our plot we want to make sure that the scale on the y-axis
will be large enough to accomodate the biggest change.
We find this with

```{r}
maxChange = max(sapply(chipsT[ -1], max))
```

### The Painter's Model

Start by making an empty plot and specify the limits of the 
y axis. Also plot the y-axis on log scale. Be sure to include
the labels and title that you created earlier.

```{r plot1}
plot( trans ~ date, data = chipsT, 
      type = "n", log = "y",
      xlab = xlabel, ylab = ylabel, main = titleP,
      ylim = c(0.5, maxChange) )
```

Use the lines() function to add three lines to the plot, one for
each measure of the chip, i.e., trans, cspd, and mps.

```{r plot2}
lines(trans ~ chips$Date, col="red",lwd=2.5))
lines(cspd ~ chips$Date, col="blue",lwd=2.5)
lines(mps ~ chips$Date, col="black",lwd=2.5)
```

### Color

To distibguish between the lines, we will use color.
Let's choose colors from one of Cindy Brewer's palettes.
To do this, follow the steps below.

* Install the RColorBrewer package. This only needs to be done once, and
do not put this code in the Rmd file.

* Load the package into your workspace with  library(RColorBrewer) 

* To see all of the colors available call the function 
*display.brewer.all()*
Call this function from the console. Do not put this call in your Rmd
file. You probably want to zoom the image so that you can see the palettes and colors better
 
* Choose a palette and colors within the palette to use. Since we are
using colors to distinguish between variable, we should choose colors 
from a qualitative/categorical palette. 
 
* Now to get the color vector to use. 
Suppose the palette has n colors, and you want the n1st, n2nd, and n3rd of them. 
Then you can call the brewer.pal as follows:
 
```{r}
library(RColorBrewer)
myColors = brewer.pal( 3, "Set2")
```

Replace the n, n1, n2, n3, and "name..." in the code above 
with the information that you want and remove the comment #. 
Then, remake the lines using your colors

```{r plot3}
lines(trans ~ chips$Date, col = myColors[1],lwd = 2.5)
lines(cspd ~ chips$Date, col= myColors[2],lwd = 2.5)
lines(mps ~ chips$Date, col= myColors[3],lwd = 2.5)
```


Now that your plot has colors, 
add a legend to the plot with the legend() function.
Update the call to legend() below to have more informative
labels and to use the colors you chose from one of Brewer's palettes.

```{r plot4}
legend("topleft", 1976, 10000, 
       legend = c("TRANS", "CSPD", "MPS"), 
       fill = myColors, bty = "n")
```


### Reference Lines 

Lastly, add  reference lines and text to the plot.
Specifically, use abline to add a vertical line at 1985 and at 1993.
This is when the first 32 bit processor and the Pentium processor
were introduced, respectively.
Also add text next to the lines to indicate
the processor that was released at that date.

With abline(), we can specify a vertical line with the v parameter,  
a horizontal line with the h parameter, or a line with any slope 
and intercept with the a and b paramters. Add two vertical lines,
one at 1983 and the other at 1985. Make them both grey and thick enough
to easily see.

```{r plot5}
abline(v = 1993, col = 'grey', lwd = 2.5)
abline(v = 1985, col = 'grey', lwd = 2.5)
  
```

Next, place text next to each line to indicate the event that
they correspond to. We can use the text() function to do this.
The text() function places text in the plotting region.
With text, you specify the x,y location of the text.

Instead, we use mtext() to place text in the margin of the plot.
We will need to specify fie parameters to the function call: 
text, side, line, at, and adj. Read the help for mtext to figure out
what these parameters do.
For the text parameter, provide a short meaningful label for the 
each vertical line.
Use the side parameter to place the text at the top of the plot.
(Recall that 1 is bottom, 2 is left, 3 is top, and 4 is the right side.)
The line parameter indicates which line of the margin to use.
This can be negative to plot within the plotting region, and it 
need not be an integer. For example, try -1.2.
The at parameter will specify the x position to place the text.
Lastly with adj you can specify where you want the text to be placed 
relative to the at position provided.

```{r}
par(oma=c(1, 1, 1, 1), mfrow=c(2, 2))
 mtext(outer=T, "The title for this page of plots", side = -1.2, line = 1, adj=0)
```


You should have a beautiful plot at this point. Congratulations!
